

```{r}
library(tidyverse)
library(dplyr)
```

```{r}
setwd(".")
data <- read.csv(file=file.path("../src/processed_data/", "Processed_Data_V1.csv"))
```

```{r}
head(data)
```

The data that we are trying to operationalize is:
- vaccination rate: `Administered_Dose1_Pop_Pct`
- trips over 100 miles: `Number.of.Trips.100.250` + `Number.of.Trips.250.500` + `Number.of.Trips...500`
- Trips over 100 miles per capita: ...
- airport access: `Num.Enplanements` (replace NA with 0...)
- education achievement: `bachelors_degree_or_higher_percent`
- income : `Per.capita.personal.income..dollars..2.`
- car access: (`Total.Households`-`No.vehicle.available`)/`Total.Households // Fraction of households with at least 1 car
- demographics: (`WA_MALE` + `WA_FEMALE`)/`TOT_POP` // Fraction of households that are white? 
- population: `TOT_POP`

```{r}
subset <- data$Administered_Dose1_Pop_Pct
subset$Num.Enplanements <- data$Num.Enplanements
subset$bachelors_degree_or_higher_percent <- data$bachelors_degree_or_higher_percent
subset$Per.capita.personal.income <- data$Per.capita.personal.income..dollars..2.

subset$Num_Trips_Over_100 <- data$Number.of.Trips.100.250 + data$Number.of.Trips.250.500 + data$Number.of.Trips...500
subset$Num_Trips_per_cap <- subset$Num_Trips_Over_100/data$TOT_POP
subset$Own_Car <- (data$Total.Households - data$No.vehicle.available) / data$Total.Households
subset$Pct_White <- (data$WA_FEMALE + data$WA_MALE) / data$TOT_POP
```

```{r}
#hist(log10(subset$Num_Trips_Over_100)) # log transform it is
hist(subset$Num_Trips_per_cap) # We will work with this 
```
```{r}
hist(log10(subset$Num.Enplanements)) # log transform on num enplanements 
hist(subset$Num.Enplanements/1000) # ???????? keep thinking about this one? 
```
```{r}
hist(subset$Pct_White) # log transform or nah - nah all good. 
```
```{r}
hist(subset$Per.capita.personal.income) # leave alone? 
```
```{r}
hist(as.numeric(subset$bachelors_degree_or_higher_percent)) # leave alone 
```