

```{r}
library(tidyverse)
library(dplyr)
```

```{r}
setwd(".")
data <- read.csv(file=file.path("./processed_data/", "Processed_Data_V1.csv"))
```

`data` is the result of the complete data merge performed in `data_processing.Rmd`. We need to down-select the variabes we will use for modeling, and perform some necessary scaling/conversions to operationalize everything. The data we are trying to model is:

- Vaccination rate
  - `Administered_Dose1_Pop_Pct` percentage of population that has received at least 1 dose as of May 15, 2021. 
- Trips over 100 miles per capita
  - (`Number.of.Trips.100.250` + `Number.of.Trips.250.500` + `Number.of.Trips...500`) / `TOT_POP` total number of trips over 100 miles taken per capita
- Airport access
  - `Num.Enplanements` total number of enplanements during the year 2017 to represent airport business
- Education achievement
  - `bachelors_degree_or_higher_percent` percentage of people in a county who hold a bachelors degree or higher
- Income
  - `Per.capita.personal.income..dollars..2.` personal income per capita
- Car access
  - (`Total.Households`-`No.vehicle.available`)/`Total.Households` fraction of households with at least one car
- Demographics
  - (`WA_MALE` + `WA_FEMALE`)/`TOT_POP` percentage of people in the county who are white

Some other necessary descriptive variables are:
- State
- County
- Total county population

```{r}
head(data)
```
```{r}
filtered_data <- data %>%
  mutate(County = County.Name.Filt.On, 
         TotPop = TOT_POP, 
         State = State, 
         GeoName= GeoName, 
         WhitePct = (WA_FEMALE + WA_MALE) / TOT_POP * 100, 
         BachelorsPct = bachelors_degree_or_higher_percent, 
         Dose1Pct = Administered_Dose1_Pop_Pct, 
         CarPct = (Total.Households-No.vehicle.available)/Total.Households * 100, 
         IncomePerCap = Per.capita.personal.income..dollars..2., 
         AirportEnplanements = Num.Enplanements, 
         TripsOver100PerCap = (Number.of.Trips.100.250 + Number.of.Trips.250.500 + Number.of.Trips...500) / TOT_POP, 
         TripsOver250PerCap = (Number.of.Trips.250.500 + Number.of.Trips...500) / TOT_POP,
         .keep = "none")
```

```{r}
head(filtered_data)
```

```{r}
write.csv(filtered_data, "./processed_data/Selected_Data_V1.csv")

```
